The MACE is essentially unidimensional: Bifactor models of the `Maltreatment and Abuse Chronology of Exposure' scale

### 2022-08-11 manuscript/main.tex
manuscript edits (add formative vs. reflective indicator discussion to supplement)

To investigate the factor structure of the MACE, we fit a series of confirmatory item factor models separately to the response data from each sample. The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each of the four models is depicted in Figure \ref{fig:models}. The first model is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b). This model will be used to evaluate if the MACE can produce reliable subscale scores for each of the 10 types of maltreatment assessed by the MACE.

@@

\begin{enumerate}
    \item I'm able to blink my eyes without difficulty. (all endorse)
    \item I was motivated to write Mumfred Mumford's biography. (none endorse)
    \item I worry about the 2001 Cricket World Cup. (none endorse)
    \item Please select "Yes" and then select age "17". (instructed item)
\end{enumerate}

@@

%One might question whether it is appropriate to model childhood maltreatment data using common factor models (e.g. bifactor models). The core assumption of these models is that the direction of causality is from the factor to the indicators. That is, variation in item responses are caused by (or reflect) variation in a shared factor. This is in contrast to composite variable models where the direction of causality is reversed; there the indicators cause (or form) the latent variable. Common factor and composite variable models are differentiable in two important ways \citep{bollen1991conventional, bollen2011three}. First, covariation among indicators is expected under a common factor model due to the influence of a shared common factor; conversely, covariation among indicators is not required (though may still occur) under a composite variable model. Second, indicators under the common factor model, but not the composite variable model, are expected to share causal antecedents because they reflect a common source. Multiple authors have argued that adverse childhood experiences should (often) be treated as formative, not reflective \citep{netland2001assessment, layne2010unpacking}. Indeed it is the occurrence of abuse and neglect that defines childhood maltreatment, not the other way around. Moreover, theoretical models of the consequences of childhood maltreatment neither require that adverse experiences covary nor assume they share causal antecedents.

%For the purposes of explaining patterns of responding on the MACE, we think we are justified in using a common factor model. The vast majority of items on the MACE measure maltreatment by parents or other adults in the home. Thus, the homogeneity of these indicators makes it plausible that they reflect a common source (e.g. parental maltreatment, household dysfunction), as does their pattern of covariation as evidenced by strong loadings onto a common factor. Moreover, there is theoretical support for a model in which a common source like parental maltreatment, itself caused by a diversity of parent level (e.g. emotional lability, impulse control, history of childhood abuse) and child-parent level factors, resulting in a multiple of forms of maltreatment \citep{belsky1993etiology}. This is empirically supported insofar that multiple forms of maltreatment share overlapping risk factors \citep{stith2009risk, mulder2018risk, assink2019risk}. Furthermore, the MACE is a retrospective self-report measure. Prospective and retrospective measures do not have perfect agreement \citep{patten2015retrospective, reuben2016lest}, and thus must reflect other factors related to the participants including individual differences in current state, memory, appraisal, personality, and biases in discloure \citep{susser2012still, reuben2016lest, colman2016consistency}.

%This in turn raises another question: whether it is reasonable for indicators of parental maltreatment and peer victimization on the MACE to share a common factor. At first blush, one might reasonably assume that the causes of the former would be independent of the causes of the latter. However, there is ample evidence that children who have been maltreated are at greater risk for subsequent peer victimization \citep{bolger2001developmental, benedini2016cycle}. The reasons for this association are manifold. For example, maltreatment may influence a child's sociocognitive development (e.g. attachment style, emotion regulation) in ways that may predispose them to peer victimization \citep{goemans2021child, McCrory2022-cj}. Parental neglect, or the absence of adult supervision, is also a risk factor for peer victimization \citep{lereya2013parenting}, potentially by increasing the likelihood of children participating in higher risk social situations where victimization may take place. Insofar that parental maltreatment is an indirect cause of peer victimization (mediated by a number of complex psychosocial pathways), then we believe we are justified specifying indicators of peer victimization to load in part onto the common factor. This is also highlights the need for future study of the peer victimization factor on the MACE, to better understand how it is related to and distinct from childhood maltreatment by parents and what its downstream consequences are.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.82\textwidth,center]{figures/figS01.png}
    \caption{\normalfont \begin{spacing}{1.25} Differential item functioning by sample. (A) Proportions of participants endorsing having experienced a maltreatment event by study sample. (B) The log odds that the proportion of endorsements are larger in one or the other sample. Dotted lines indicate the threshold for large DIF recommended by \cite{hidalgo2014binary}. ** Reverse-scored items. \end{spacing}}
    \label{fig:dif_study}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.82\textwidth,center]{figures/figS02.png}
    \caption{\normalfont \begin{spacing}{1.25} Differential item functioning by gender. (A) Proportions of participants endorsing having experienced a maltreatment event by study gender. (B) The log odds that the proportion of endorsements are larger in one or the other gender. Dotted lines indicate the threshold for large DIF recommended by \cite{hidalgo2014binary}. ** Reverse-scored items. \end{spacing}}
    \label{fig:dif_gender}
\end{figure}

>>>>

To investigate the factor structure of the MACE, we fit a series of confirmatory item factor models\footnote{The appropriateness of common factor models with reflective indicators for maltreatment data has been contested. We justify our usage of this model of the MACE for the interested reader in the supplement.} separately to the response data from each sample. The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each of the four models is depicted in Figure \ref{fig:models}. The first model is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b). This model will be used to evaluate if the MACE can produce reliable subscale scores for each of the 10 types of maltreatment assessed by the MACE.

@@

\begin{enumerate}
    \item I'm able to blink my eyes without difficulty. (all endorse)
    \item I was motivated to write Mumfred Mumford's biography. (none endorse)
    \item I worry about the 2001 Cricket World Cup. (none endorse)
    \item Please select ``Yes'' and then select age ``17''. (instructed item)
\end{enumerate}

@@

\break
\subsection*{Reflective vs. formative indicator models of maltreatment data}

One might question whether it is appropriate to model childhood maltreatment data using common factor models with reflective indicators (e.g. bifactor models). The core assumption of these models is that variation in item responses are caused by or reflect variation in a shared latent factor (e.g. variation in low mood and anergia reflect latent variation in depression). This is in contrast to composite variable models with formative indicators where the direction of causality is reversed \citepSM{bollen1991conventional}. In composite variable models, indicators cause or form the latent construct (e.g. education and income together make up socioeconomic status). Some scholars have argued that measures of adverse childhood experiences often should be treated as formative, not reflective, indicators \citepSM{netland2001assessment, layne2010unpacking}. That is, exposure to abuse and neglect defines childhood maltreatment, not the other way around.

For the purposes of explaining covariation in maltreatment endorsements on the MACE, we believe we are justified in using a common factor model. Most of the items on the MACE measure maltreatment by parents or other guardians. The homogeneity of these indicators suggests they might reflect a common source; this hypothesis is further evidenced by the uniformly strong general factor loadings for this subset of items. Furthermore, there is ample theoretical and empirical support for regarding parental maltreatment as a common factor with its own causal antecedents \citepSM{belsky1993etiology}. Indeed, many of the same parent-level variables (e.g. emotional lability, impulse control, own history of abuse) are risk factors for different types of childhood maltreatment \citepSM{stith2009risk, mulder2018risk, assink2019risk}. Separately, that the MACE is a retrospective self-report measure is further cause for a common factor model. Retrospective measures of abuse have been shown to reflect additional factors beyond maltreatment exposure including individual differences in current emotional state, personality, and recall ability \citepSM{susser2012still, reuben2016lest, colman2016consistency}. Common factor models may better adjust for some of the measurement error in indicators caused by these nuisance factors.

This discussion in turn raises a second question: whether it is reasonable for indicators of parental maltreatment and peer victimization on the MACE to share a common factor. One might assume that these two types of maltreatment reflect independent causes, and thus should not be permitted to share a common factor. There is considerable evidence, however, that children who have been maltreated are at greater risk for subsequent peer victimization \citepSM{bolger2001developmental, benedini2016cycle}. The reasons for this association are manifold. For example, parental maltreatment may impact a child's sociocognitive functioning (e.g. attachment styles, emotion regulation) in ways that may predispose them to later peer victimization \citepSM{goemans2021child, McCrory2022-cj}. Separately, parental neglect is a risk factor for peer victimization \citepSM{lereya2013parenting}, possibly by increasing the chance that children enter into risky social situations where victimization may occur. Insofar that parental maltreatment is an indirect cause of peer victimization (mediated by multiple complex psychosocial pathways), then we believe we are justified in allowing indicators of peer victimization to load onto the common general maltreatment factor. This is also highlights the need for future study of a MACE peer victimization subscale to better understand how it is related to (and distinct from) parental maltreatment, and what its downstream consequences are.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.82\textwidth,center]{figures/figS01.png}
    \caption{\normalfont Differential item functioning by sample. (A) Proportions of participants endorsing having experienced a maltreatment event by study sample. (B) The log odds that the proportion of endorsements are larger in one or the other sample. Dotted lines indicate the threshold for large DIF recommended by \cite{hidalgo2014binary}. ** Reverse-scored items.}
    \label{fig:dif_study}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.82\textwidth,center]{figures/figS02.png}
    \caption{\normalfont Differential item functioning by gender. (A) Proportions of participants endorsing having experienced a maltreatment event by study gender. (B) The log odds that the proportion of endorsements are larger in one or the other gender. Dotted lines indicate the threshold for large DIF recommended by \cite{hidalgo2014binary}. ** Reverse-scored items.}
    \label{fig:dif_gender}
\end{figure}
